---
title: "SQL exercise - basics - SOLUTION"
description: |
  Very basic exercises for SQL in case you still need practice.
author:
  - name: Mirko Mälicke
    url: https://hyd.iwg.kit.edu/personen_maelicke.php
    affiliation: Karlsruhe Institute for Technology (KIT)
    affiliation_url: https://hyd.iwg.kit.edu
date: "`r Sys.Date()`"
output:  
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
  radix::radix_article:
    toc: yes
  pdf_document:
    toc: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# require the package
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
if (!("getPass" %in% installed.packages())){
  install.packages("getPass")
}
require(RPostgreSQL)
require(getPass)
# establish the connection
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='v45522.1blu.de', port=5432, user=getPass('Provide the user'), 
                 password=getPass('Provide the password'), dbname='datamanagement')
```

# Arithmetics, casts

How many records can be found in the raw_data table?
```{sql connection=con}

```

How many hobo entires are in the hobo table:
```{sql connection=con}

```

How many distinct hobo sensors have been used?
```{sql connection=con}

```

You can also cast data types in a <span style="color: blue;">SELECT</span> and do some basic math. Use this to query the percentage of reused HOBOs
```{sql connection=con}

```

Cast the result from above into a string and use the concatenate operator: ``||`` to create a proper message
```{sql connection=con}

```

# filters

In which of the last three year participated the most students?
```{sql connection=con}

```

You want to find the maximum light intensity record in the raw_data table. What is faster, aggregating the whole column using the max function, as you might do this in R, or order the table by light and take only the first row. Implement both and <span style="color: blue;">EXPLAIN ANALYZE</span> both.
```{sql connection=con}
select hobo_id, tstamp, light from ...
```
```{sql connection=con}
select hobo_id, tstamp, light from ...
```
```{sql connection=con}
explain analyze 
```
```{sql connection=con}
explain analyze 
```

This is not that that much faster as you might have expected. One advantage of the aggregation solution is that you can find out wether there is more than one value that fits this constraint. This is not the case if you limit the query by 1. 


# cleanup
```{r}
dbDisconnect(con)
```

