---
title: "PostGIS"
description: "Enable powerful GIS functionality in your PostgreSQL database."
author:
  - name: "Mirko Mälicke"
    url: "https://hyd.iwg.kit.edu/personen_maelicke.php"
    affiliation: "Karlsruhe Institute for Technology (KIT)"
    affiliation_url: "https://hyd.iwg.kit.edu"
date: "`r Sys.Date()`"
output:  
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
  radix::radix_article:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
# require the package
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
if (!("getPass" %in% installed.packages())){
  install.packages("getPass")
}
if (!("plotly" %in% installed.packages())){
  install.packages("plotly")
}
if (!("sf" %in% installed.packages())){
  install.packages("sf")
}
library(RPostgreSQL)
library(getPass)
library(plotly)
library(sf)

pw <- function () {
  if (Sys.getenv('POSTGRES_PASSWORD') == ""){
    return(getPass('Provide the password: '))
  } else {
    return(Sys.getenv('POSTGRES_PASSWORD'))
  }
}

# establish the connection
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='hydenv.hydrocode.de', port=5432, user='hydenv', 
                 password=pw(), dbname='hydenv')
```

# PostGIS

PostGIS is a PostgreSQL extension that offers several new datatypes for attribute declaration along with hundreds of predefined functions.
The main new data type is called `GEOMETRY` and can store any kind of OGR defined geometry object. 
The most important ones are the `'POINT'`, `'LINESTRING'` and `'POLYGON'` geometry. A new table `spatial_ref_sys` is also created, which includes almost any existing CRS definitions. 
The functions can be used for any kind of spatial and geometrical operation and follows common naming conventions you might already know from other GIS systems. In fact, the PostgreSQL / PostGIS system is a full featured GIS system. <br>
CRS are stored in an ordinary table and can be queried just like you did it in the other lectures:

```{sql connection=con}
select * from spatial_ref_sys limit 5
```

This looks quite simple. There is a srid, the primary key, which you should already know. The creators of PostGIS used the EPSG number 
as primary key. This is very handy as the ESPG identifiers are already unique (and well known).<br>
The _proj4text_ field stores the CRS definition as a Poj4 string. This can be understood by most GIS system, like QGis, and recreate 
missing or custom CRS. The _srtext_ field stores the WKT (well known text) definition of the CRS. This could be used to build the CRS 
by hand in other programming languages like Python, Java or C++. <br>
Ok, then let's see if all coordinate systems we might need are there:
<ul>
<li>unprojected WGS84, as kown from google maps</li>
<li>Pseudo Mercartor, as used by openstreetmap</li>
<li>DHDN, Gauß Krüger Z3, the old offical CRS in Baden-Württemberg</li>
<li>ETRS89 / UTM Z32N, new new official CRS in Baden-Württemberg</li>
</ul>

```{sql connection=con}

```

<div class="alert alert-info">Usually, you don't have to interact with the spatial reference systems directly as PostGIS will take care of that. Nevertheless it's good to know, where it finds its informations to do some magic.</div>

# Loading spatial data to R

You might have recognized, that RStudio is complaining about a not known datatype when querying the stations table from the database. This is due to the GEOMETRY data type, we introduced earlier. This is not known by the R package RPostgreSQL. Unlike Python, there is no easy way to make R or RStudio understand this datatype. Therefore, we will always have to load the geometries in the WKT format and convert it in R back to a binary spatial object as supported by the R packages you prefer. The alternative approach is to use the simple features `sf` package. That is a great package, but it has some nasty dependencies, that do not easily interact with other software on a productive system. We will cover both approaches here.<br>
This is where the PostGIS funcitons set in. We can use any of them in <span style="color:blue">SELECT</span> statements or 
<span style="color:blue">WHERE</span> filters. In case you are familiar with the GDAL C++ or Python package, GRASS GIS, Saga GIS, whitebox-gis or any other command line based GIS solution, most of the PostGIS functions will be quite familiar to you. Otherwise you will have to search the documentation for the correct function names and usages.<br>
The WKT of any GEOMETRY object can be loaded with the *ST_AsEWKT* function:

```{sql connection=con}

```

We can see two things here: First, we are able to read the coordinates now. Second, the raw WKT geometry information is prefixed by an information on the used CRS. This is a special PostGIS definition that might not be understood by every other GIS system. The advantage is the ability of the system to store the geometry and CRS information in only one attribute. This also means, that there is no need to connect this table to the spatial_ref_sys table anymore. We could transform these coordinates on select.
You can select a location twice, ie. in WGS84 and projected UTM representation.

```{sql connection=con}

```

From here, we have several options to put the geometry information into a more usable format for R. We either load a package that can read WKT. Secondly, we could parse the Strings and extract the needed information ourselves or we could query the information in a more readable format for R. 

<div class="alert alert-warning">Other languages like Python offer way more powerful bindings to databases. The SQLAlchemy and geoalchemy2 packages in Python, or the pandas/geopandas/polars data-science packages, which are available in most scientific Python environments, can load, plot and manipulate PostGIS geometries out of the box.</div>

### Base R

First, let's load some point geometries with base r.
```{sql connection=con, output.var="hobo"}
select metadata.id, device_id, short as term, as lon, as lat from metadata 
join terms on terms.id=metadata.term_id
where location is not null
```
```{r}
plot_ly(data=hobo, x=~lon, y=~lat, type="scatter", mode="markers", marker=list(size=8), 
        color=~term, text=~paste("Device ID: ", device_id)) %>% 
  layout(xaxis=list(zerolines=F), yaxis=list(zerolines=F))
```
```{r}

```


## using simple features

If you want to read other geometries, you need to handle the WKT first. Here, simple features comes in quite helpful, as it can read PostGIS geometries. We have some OSM data on the server, let's check that one out. Working with OSM data usually needs some deep joining and a fair knowledge of the OSM tag system. Here, the most important information was extracted and added to the `osm_nodes` table:
<ul>
  <li>The geometries are already dissolved</li>
  <li>The name is extracted and added</li>
  <li>For relations containing the tag `border=administrative` the tag `admin_level=9` and `admin_level=10` were translated to the node type 'district'</li>
</ul>
```{sql connection=con}
SELECT * FROM osm_nodes 
```

```{r}
districts <- st_read(con, query="SELECT name, geom as geometry FROM osm_nodes WHERE node_type='district'", quiet=T)
plot_ly(districts)
```
```{r}
plot_ly() %>% 
  add_sf(data=districts, text=~name, type="scattermapbox", mode="lines", fill="toself", color="steelblue",
         line=list(color="black"), name="City districts") %>%
  
  add_trace(data=hobo, x=~lon, y=~lat, mode="markers", type="scattermapbox",
            marker=list(size=12), color=~term, text=~paste("Device ID: ", device_id), name=~term) %>%
  
    layout(
      mapbox=list(style="stamen-terrain", zoom=10., center=list(lon=7.8, lat=48), pitch=45), 
      legend=list(orientation='h'))
```



# Enabling spatial functions

## Transformations
One of the important spatial functions is PostGIS ability to transform coordinates from and into any coordinate system defined in the spatial_ref_sys table. This is extremely helpful for combining data from different datasources, or when a target application requires a specific CRS (like web mapping libraries). Secondly, when applying spatial functions and calculating relations, distances or areas we must not use an unsuitable CRS in order to prevent big calculation mistakes.<br>
PostGIS knows two helpful functions: *ST_Transform* for applying a transformation and *ST_SetCRS* for setting the CRS information in case they are missing in the GEOMETRY object.

```{sql connection=con}

```

The difference to `ST_SetSRID` becomes obvious with the next example:
```{sql connection=con}
SELECT device_id, 
  st_asewkt(st_transform(location, 25832)) as "st_transform",
  st_asewkt(st_setsrid(location, 25832)) as "st_setsrid"
from metadata
LIMIT 10
```

## GIS operations

It is also possible to calculate distances in the database. Once calculated, these distance can be used like any other attribute, this means you can also sort or group by distances. Knowing this, it is for example easy to find the next station for a specific location, or a set of locations.<br>
Let's find a random hobo (id 42) and search the database for the the closest and furthest other device.

```{sql connection=con}
select * from metadata where id=42
```
```{sql connection=con}

```
As you can see, the distance seems to be incorrect. What is going on?
Fix this query and union this query two times, just just ordered ascending and descending.
```{sql connection=con}

```

It is also possible to subset the table and filter the stations to be within a specific distance to our random HOBO (e.g. 2.5km). To make the queries a bit shorter, let's create a temporary view, that does the transformations for us.
```{sql connection=con}
create temporary view hobo as
select st_transform(location, 25832) as geom, * from metadata
```


```{sql connection=con}
select id, device_id from hobo where st_distance(geom, (select geom from hobo where id=42)) <= 2500
```
```{sql connection=con}
select id, device_id from hobo where st_within(geom, st_buffer((select geom from hobo where id=42), 2500))
```

These two solutions lead to an identical solution, but there is something very different happening. It is very good to have different ways of calculating the same result. The way PostgreSQL will find the selected features can be described as a query plan. This is essentially, what the computer is planning to do and, more important, why it came to these decisions. Therefore it is a very helpful tool to be able to print out these query plans when trying to identify dropping performances. 
Another helpful information of a query plan is the total runtime on the machine. This does not include the time the data needs to be transferred to your computer and the time a client application needs to visualize the data, which is usually the bottleneck but has nothing to do with the actual database performance.<br>
In PostgreSQL you can prefix any <span style="color:blue">SELECT</span> statement with <span style="color:blue">EXPLAIN ANALYZE</span>to make PostgreSQL print out the query plan instead of the results.

```{sql connection=con}
explain analyze select id, device_id from hobo where st_distance(geom, (select geom from hobo where id=42)) <= 2500
```
```{sql connection=con}
explain analyze select id, device_id from hobo where st_within(geom, st_buffer((select geom from hobo where id=42), 2500))
```

# GIS in the database

## Working with different geometries

The last example for this lecture is a GIS analysis. We will switch back to the CAMELS-DE dataset. Along with the discharge and water level data, we have metadata, which contains location information.
```{sql connection=con, output.var="sh"}
SELECT *, 
  st_x(st_transform(geometry, 4326)) as lon,
  st_y(st_transform(geometry, 4326)) as lat
from camels_de_metadata where q_count > 0
```
```{r}
sh %>%
  plot_ly(type="scattermapbox", mode="markers", lon=~lon, lat=~lat, marker=list(size=10, color="brown", opacity=0.9)) %>%
  layout(mapbox=list(style="open-street-map", zoom=6.8, center=list(lon=10, lat=54.15)))
```

In addition there is river data from openstreetmap in the database. It is already substantially pre-processed in a way that you can simply filter for the `node_type='waterway'`

```{r}
waterways <- read_sf(con, query="SELECT id, name, geom as geometry from osm_nodes WHERE node_type='waterway' and name is not null and name != 'Indian River'", quiet=T)
```
```{r}
library(ggplot2)
ggplotly(ggplot() + geom_sf(data=waterways, aes(color=name)))
```

Although the plot looks good, the hover is snapping to many features. A reason for this could be, that the rivers consist of different parts. This can easily be verified.
```{sql connection=con}
SELECT name, count(*) from osm_nodes where node_type='waterway' and name is not null 
group by name
```

Now, we can use a geometric operation instead of the `count` function for the aggregation:
```{sql connection=con}
drop table if exists rivers;
create temporary table rivers as


```
```{sql connection=con}
select name, round(st_length(st_transform(geom, 3035)) / 1000) as "length [km]" from rivers order by "length [km]" DESC
```

Next, we could join each discharge station to the next waterway, but to practice a bit more PostGIS, we will aggregate the discharge stations and count them, if they are closer than 15 meters to the stream, by buffering the river.
```{sql connection=con}
with buffers as (
  select name, st_buffer(st_transform(geom, 3035), 15) as buffer from rivers
),
stations as (
  select name, count(camels_id) as n_stations from buffers
  join camels_de_metadata c on st_within(c.geometry, buffers.buffer)
  group by name
)
select r.name, round(st_length(st_transform(geom, 3035)) / 1000) as "length [km]", s.n_stations
from rivers r
join stations s on s.name=r.name
order by n_stations DESC
```

And finally, we can aggregate the data along with the aggregation of the stations. If this is very useful from a hydrological perspective is up to you, but remember that much more powerful aggregations, than just an average value exist. We can do some simplified extreme statistics.
For this we first need to take care of the discharge data and use some partitions, so that we can preserve the date at which the maximum discharge was observed. If we would simply group by year and aggregate with the `max` function, we lose that information.
```{sql connection=con}
drop table if exists year_max;



```
```{sql connection=con}
select * from extremes limit 100
```

Now, we can aggregate the table created above, which includes the return interval for every recorded year to only those with an empirical return interval larger than 10 years. We will use an array aggregation, to collect all values.

```{sql connection=con}
select camels_id, array_agg("T"::integer) as "T", array_agg(q) as q, array_agg(date) as date
from extremes
where "T" > 10
group by camels_id
```

Finally, we can combine both, the discharge stations for each river segment and the extreme statistics. The workflow is to calculate the statistics per station, then join them to a river segment and finally aggregate the river segments by their name to have extreme statistics per river.

For this, it is useful to create a temporary table containing the merged extreme statistics and the river name. For this, we need to make a spatial join that merges the river name based on a 15 meter buffer around the rivers to the contained stations. This might seem to be a bit more complicated, but it will be easier to perform the final aggregation.
```{sql connection=con}
drop table if exists stats_with_rivers;
create temporary table stats_with_rivers as



```
```{sql connection=con}
select * from stats_with_rivers limit 10;
```
Now we have each CAMELS-DE station with all return periods, discharges and dates of flood with an empirical return period larger than 10 years. Finally, we can do some analysis on this dataset.

Create statistics for each river:
```{sql connection=con, output.var="extremes"}


```
```{r}
extremes
```

Now, using nested data is really wild in R. It would be a good idea to use a more appropriate language here like Python or Typescript.
```{r}

```

# cleanup
```{r}
dbDisconnect(con)
```





