---
title: "SQL exercise - advanced - SOLUTION"
description: |
  Intermediate SQL exsercise to make you familiar with powerful database concepts.
author:
  - name: Mirko Mälicke
    url: https://hyd.iwg.kit.edu/personen_maelicke.php
    affiliation: Karlsruhe Institute for Technology (KIT)
    affiliation_url: https://hyd.iwg.kit.edu
date: "`r Sys.Date()`"
output:  
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
  radix::radix_article:
    toc: yes
  pdf_document:
    toc: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# require the package
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
if (!("getPass" %in% installed.packages())){
  install.packages("getPass")
}
require(RPostgreSQL)
require(getPass)
# establish the connection
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='v45522.1blu.de', port=5432, user=getPass('Provide the user'), 
                 password=getPass('Provide the password'), dbname='datamanagement')
```

# Palying with Indices

There is data of three student generations in the database. Which one was the warmest? Which one the coldest year in average?
```{sql connection=con}

```

How do these temperatures look like if you exclude 30% of the brightest measurements as they might be considered as highly influenced by radiation.
```{sql connection=con}

```

And additionally the 10% of warmest measurements?
```{sql connection=con}

```

Calculate the minimum temperature for each day, when the light sensor wasn't recording any radiation across all HOBO sensors of 2019
```{sql connection=con}

```

Order the last result by ascending temperatures and subset only the coldest 40%. What is their average minimum daily-dark temperature?
```{sql connection=con}

```


What did the same index look like for 2017 and 2018, complete the task in one query.
```{sql connection=con}

```

Take again the coldest 40% of the daily minimum temperatures of these generation recordings, while it was dark. How many percent were observed in 2018, how many in 2019? 
```{sql connection=con}

```



# cleanup
```{r}
dbDisconnect(con)
```