---
title: "SQL exercise - PostGIS - SOLUTION"
description: |
  PostGIS focused exercise to get familir with GUI-free GIS.
author:
  - name: Mirko Mälicke
    url: https://hyd.iwg.kit.edu/personen_maelicke.php
    affiliation: Karlsruhe Institute for Technology (KIT)
    affiliation_url: https://hyd.iwg.kit.edu
date: "`r Sys.Date()`"
output:  
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
  radix::radix_article:
    toc: yes
  pdf_document:
    toc: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# require the package
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
if (!("getPass" %in% installed.packages())){
  install.packages("getPass")
}
require(RPostgreSQL)
require(getPass)
# establish the connection
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='v45522.1blu.de', port=5432, user=getPass('Provide the user'), 
                 password=getPass('Provide the password'), dbname='datamanagement')
```

# GIS

You already know the catchments table in the database. One important first analysis is whether the features overlap or not. There are many ways how we could accomplish this task. One possibility is to compare the total sum of feature areas to the area of the union.
How big is the total covering area of the catchments are? How big is the total area of all features?
```{sql connection=con}

```
```{sql connection=con}

```

Find the catchment containing most of this years HOBO locations
```{sql connection=con}

```

Use the geometry of the catchment identified in the last exercise to find the evapotranspiration (from the same named table) for exactly this area. By how many features is this area represented? Load the features.
```{sql connection=con}

```

What is the mean evapotranspiration, based on the last query?
```{sql connection=con}

```

Remember, that we just loaded the featues that intersected with the geometry of the catchment. This does not mean, that all the features lie completely in the catchment area. Now, recalculate the average annual evapotranspiration as an area weighted mean value, using only the area actually contained by the catchment geometry. You might at first query an overview of the corrected geometry areas along with their mean ET (attribute 'mitte'): 
```{sql connection=con}

```
Then calculate the weighted mean from this:

```{sql connection=con}

```

Compare this ET to the weighted mean ET if you would forget the intersection of the correct areas
```{sql connection=con}

```


# cleanup
```{r}
dbDisconnect(con)
```

